% Compile with LuaLaTeX: `lualatex writeup_updown.tex`
\documentclass[11pt, a4paper]{article}
\usepackage{fontspec}
\usepackage[english]{babel}
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\def \N{{\mathbb Z}_{\geq 0}}
\def \Npos{{\mathbb Z}_{>0}}
\def \Z{\mathbb Z}
\linespread{1.2}
\setlength{\parindent}{0em}

\begin{document}

\section{Setup}
Let $S$ be a numerical semigroup, with $f,\,m,\,g,\,\sigma$ as usual
(Frobenius, multiplicity, genus, number of sporadic elements).
Write $S_+ := S \setminus \{0\}$.
\begin{itemize}
    \item $\operatorname{Ap}(S):=\{s\in S \mid s-m \notin S\}$, the
    \emph{Apéry set} (one representative per residue class mod $m$).
    Index it by residue: $\operatorname{Ap}(S) = \{w_0,\dots,w_{m-1}\}$
    with $w_i \equiv i \pmod m$. In particular $w_0 = 0$.
    
    \item $\operatorname{RG}(S) := \{L \in \{1,\dots,f-1\}
    \mid L \notin S,\ f-L \notin S\}$, the \emph{reflected gaps}.
    
    \item $a$ is an \emph{atom} of $S$ iff
    $a \in S_+ \setminus (S_+ + S_+)$. Equivalently, $a$ lies in the
    unique minimal generating set $\operatorname{Atom}(S)$ of $S$.
    \item $\mu \in \{1,\dots,m-1\}$ denotes $\mu := f \bmod m$;
    equivalently $f+m = w_\mu$ (the largest Apéry element).
    \item For each $i \in \{1,\dots,m-1\}$, set
    \[
        r_i := \frac{w_i + w_{\mu-i} - w_\mu}{m},
    \]
    where the index $\mu - i$ is taken mod $m$. Geometrically:
    $w_i + w_{\mu-i}$ sits exactly $r_i$ steps of $m$ above $w_\mu$
    on the residue-$\mu$ ladder, so the closed interval
    $[w_\mu, w_i + w_{\mu-i}]$ in that ladder contains $r_i + 1$ rungs
    (the rung $w_\mu - m = f$ being a gap and the rest of the rungs
    above $w_\mu$ being elements of~$S$). $r_i$ counts the reflected
    gaps in residue class $i \pmod m$.
    
    \item $V(S,l) := \{f-lm+1,\,f-(l-1)m+2,\dots,f-1\}$, the $l$-Stratum, for 
    $l=1$ this is the 
    \emph{interval below $f$} (always $m-1$ consecutive integers).
    \item $\rho(S) := \min_{i \in \{1,\dots,m-1\}\setminus\{\mu\}} r_i$.
    The class $i = \mu$ is excluded because $r_\mu = 0$ trivially
    (every $x \equiv \mu$ has $f - x \equiv 0 \pmod m$, hence
    $f - x \in S$).
    \item Total reflected-gap count:
    $r = \sum_{i=1}^{m-1} r_i = \#\operatorname{RG}(S)$
    (the term $r_0$ would equal $0$ anyway).
    \item Trivial identities: $r + \sigma = g$, $f + 1 + r = 2g$,
    and $t - 1 \le r$. The first follows from closure: no pair
    $\{x, f-x\}$ can lie inside $S_+$ (else $f \in S$), and similarly
    $f/2$ is always a gap when $f$ is even; pairing up
    $\{0,1,\dots,f\}$ under $x \leftrightarrow f-x$ then gives
    $r + \sigma = g$. The second is $g + \sigma = f+1$ combined with
    the first. The third holds because each pseudo-Frobenius
    $z \neq f$ is a reflected gap ($z + S_+ \subseteq S$ would imply
    $f \in S$ if $f - z \in S_+$).
\end{itemize}

\section{UpDown}

Fix $m > 2$. We define four families of one-step moves, two
\emph{descents} ($S \mapsto S^\downarrow$, adding one element) and
two \emph{ascents} ($S \mapsto S^\uparrow$, removing one atom).
Primes denote the new counts in $S^\downarrow$ resp.\ $S^\uparrow$.
The descents are exactly the inverses of the ascents
($\mathcal{A}_{flip}$ inverts $\mathcal{D}_{flip}$, with
$w_i \in V(S)$ playing the role of $w_i - m$;
$\mathcal{A}_{total}$ inverts $\mathcal{D}_{total}$).

\begin{enumerate}
    \item $\mathcal{D}_{flip}(i)$, with $i \in \{1,\dots,m-1\}\setminus\{\mu\}$
    and $f < w_i < f + m$:
    \[S^\downarrow := S \cup \{w_i - m\}, \qquad
       g' = g - 1,\ \sigma' = \sigma + 1,\ r' = r - 2.\]

    \item $\mathcal{D}_{total}(\mu)$, requiring $V(S) \subseteq S$:
    \[S^\downarrow := S \cup \{w_\mu - m\} = S \cup \{f\}, \qquad
       g' = g - 1,\ \sigma' = \sigma - (m-1),\ r' = r + (m-2).\]

    \item $\mathcal{A}_{flip}(i)$, with $i \in \{1,\dots,m-1\}\setminus\{\mu\}$,
    $ka=w_i$ and Apery--Element for an atom a of $S$ and $k>0$, and $w_i \in V(S,l)$ (going down on l if necessary):
    \[S^\uparrow := S \setminus \{a\}, \qquad
       g' = g + 1,\ \sigma' = \sigma - 1,\ r' = r + 2.\]

    \item $\mathcal{A}_{total}(\mu)$, requiring $w_\mu = f+m$
    to be an atom of $S$ (equivalently $\rho(S) > 0$); set
    $\rho := \rho(S)$:
    \[S^\uparrow := \bigl\langle A(S) \setminus \{w_\mu\} \bigr\rangle, \qquad
       g' = g + \rho,\ \sigma' = \sigma + \rho(m-1),\ r' = r - \rho(m-2).\]
    Note: $S^\uparrow$ is \emph{not} the set $S \setminus \{w_\mu\}$.
    The latter happens to be closed under $+$ already (when $\rho > 0$)
    but only drops one element; the structural ascent removes
    $w_\mu, w_\mu + m, \dots, w_\mu + (\rho-1)m$ in one shot, since
    those rungs of the residue-$\mu$ ladder were reachable in $S$
    only via $w_\mu$. The first rung of class $\mu$ that survives is
    $w_\mu + \rho m = w_i + w_{\mu-i}$ for some
    $i \in \{1,\dots,m-1\}\setminus\{\mu\}$.
\end{enumerate}

All four claims are proved by tracking the Apéry set under the move.
Aside: the Wilf conjecture in the form $(e-2)\sigma \ge r$ (which is
equivalent to the standard $e\sigma \ge f+1$ via $r + \sigma = g$ and
$g + \sigma = f+1$) constrains $\sigma$ and $r$ in tandem. We know
the change of both in every case, so working out the change of $e$
would close the loop.

\section{Correspondence}

We now zoom in on the total cases. Fix $m > 2$ throughout this section.

\textbf{Equivalence: $w_\mu$ atom $\Leftrightarrow \rho(S) > 0$.}
If $\rho > 0$ then $w_i + w_{\mu-i} = w_\mu + r_i m > w_\mu$ for every
$i \in \{1,\dots,m-1\}\setminus\{\mu\}$, so $w_\mu$ is not a sum of
two positive Apéry elements, hence (since every element of $S_+$ is
$w_j + km$ for some $j$ and $k \ge 0$) not a sum of two elements of
$S_+$ at all; that is, $w_\mu$ is an atom. Conversely, if some
$r_i = 0$ then $w_i + w_{\mu-i} = w_\mu$ with both summands in $S_+$,
so $w_\mu$ is decomposable.

\textbf{When can $\mathcal{D}_{total}$ and $\mathcal{A}_{total}$ undo
each other?} Suppose we start at $S$ with $\rho(S) = 0$, so
$w_\mu(S)$ is not an atom and we cannot apply $\mathcal{A}_{total}$
to $S$ directly. Apply $\mathcal{D}_{total}$ instead (assuming
$V(S) \subseteq S$): we obtain $S^\downarrow := S \cup \{f\}$, whose
Frobenius drops and whose new $w_\mu$ is the atom $f$ (a fresh
minimal generator). If the new $\rho(S^\downarrow) = 1$, then
applying $\mathcal{A}_{total}$ to $S^\downarrow$ removes exactly one
element --- the new $w_\mu$, namely the original $f$ --- and returns
us to $S$. This is the one situation in which a single descent
followed by a single ascent is a round trip:
\[
   S_{\rho = 0}
   \;\xrightarrow{\;\mathcal{D}_{total}\;}\;
   D(S)_{\rho = 1}
   \;\xrightarrow{\;\mathcal{A}_{total}\;}\;
   S.
\]
For $\rho(D(S)) > 1$ the round trip is no longer one-step in the
ascent: $\mathcal{A}_{total}$ deletes $\rho$ elements at once, so to
return to $S$ we would need $\rho$ separate descents (each
peeling off one element). Concretely, undoing the ascent's
$(g + \rho,\ \sigma + \rho(m-1),\ r - \rho(m-2))$ by $\rho$
applications of $\mathcal{D}_{total}$ gives a net change of
\[
   g + \rho - \rho \cdot 1 = g,\qquad
   \sigma + \rho(m-1) - \rho(m-1) = \sigma,\qquad
   r - \rho(m-2) + \rho(m-2) = r,
\]
as required. But each individual descent step demands
$V \subseteq S$ at that point, which may fail before the round trip
completes; see the next subsection.

\textbf{The map $D$.} Let $\mathbb{S}_\mu^{\mathrm{old}}$ be the set
of numerical semigroups $S$ with $m(S) = m$ and $\mu(S) = \mu$, and
satisfying $f(S) > m$ and $V(S) \subseteq S$. Define
\[
   D : \mathbb{S}_\mu^{\mathrm{old}} \longrightarrow
   \{\text{numerical semigroups with } m(S) = m\},
   \qquad D(S) := S \cup \{f(S)\}.
\]
(We do \emph{not} claim $D(S)$ has Frobenius congruent to $\mu \bmod m$;
in general $\mu(D(S)) \neq \mu(S)$. The subscript ``$\mu$'' refers
to the residue of $f$ in $S$, not in $D(S)$.) The following
properties are useful and are proved by tracking the Apéry set
through the descent:
\begin{itemize}
   \item $\rho(D(S)) = \rho(S) + 1$, where on the left $\rho$ is
   computed with respect to $\mu(D(S))$.
   \item In residue class $i$ relative to the new $\mu(D(S))$,
   $r_i(D(S)) = r_i(S) + 1$.
   \item $\rho(D(S)) > 0$, so $w_\mu(D(S)) = f(S) + m$ is an atom
   of $D(S)$, i.e.\ $D(S)$ is eligible for $\mathcal{A}_{total}$.
\end{itemize}
What remains to control along an iterated descent is whether
$V(D(S)) \subseteq D(S)$ --- the condition needed to apply
$\mathcal{D}_{total}$ a second time.

\subsection{One-step bijection}
The only way a single $(\mathcal{D}_{total}, \mathcal{A}_{total})$
pair is a round trip is
\[
   S_{\rho = 0}\;\longleftrightarrow\;D(S)_{\rho = 1},
\]
i.e.\ when $w_\mu(S)$ is \emph{not} an atom of $S$ but $D(S)$ has
$\rho = 1$. Only then is $D(S) \in \mathcal{A}_{total}(\mu(D(S)))$
and the ascent's removal-of-$\rho$-elements coincides with the
descent's addition of $f$.

Symmetric semigroups ($r = 0$) are excluded from
$\mathcal{A}_{total}$: $r = 0$ forces every $r_i = 0$, hence
$\rho = 0$, hence $w_\mu$ is not an atom.

\subsection{$f+m$ atom: iterated descent, $\rho(S) > 0$}
As long as $V(S) \subseteq S$ we can keep descending:
\[
   S_\rho \;\to\; D(S)_{\rho+1} \;\to\; D^2(S)_{\rho+2} \;\to\; \dots
\]
At every step $f + m$ is the new atom $w_\mu$ (so $\mathcal{A}_{total}$
remains applicable), and $\rho$ increases by $1$. The chain
terminates exactly when either $V(S) \not\subseteq S$ at the
current $S$ (so $\mathcal{D}_{total}$ no longer applies) or
$f < m$ (so $V(S)$ is empty/degenerate).



\section{Code}
\begin{verbatim}
fn test_up_downs(s: &Semigroup) {
    if s.m == 1 {
        return;
    }
    for i in 1..s.m {
        let w = s.apery_set[i];

        // Descent: w-m is a gap; adding it lifts the membership of w-m.
        if w > s.f && s.f > s.m {
            assert!(!s.element(w - s.m));
            assert!(w <= s.f + s.m);
            assert!(
                w == s.f + s.m || s.is_reflected_gap(w - s.m),
                "reflected {w} at {i}"
            );
            let down = s.toggle(w - s.m);
            assert_eq!(down.g + 1, s.g);
            if w == s.f + s.m && s.v_in_s() {
                assert_eq!(down.r + 2, s.r + s.m, "r change w==f+m for i={i}");
                assert_eq!(down.sigma + s.m, s.sigma + 1, "sigma");
            }
            if w < s.f + s.m {
                assert_eq!(down.r + 2, s.r, "r change w<f+m for i={i}");
                assert_eq!(down.sigma, s.sigma + 1, "sigma");
            }
        }

        // Ascent: w must already be a minimal generator.
        if s.gen_set.contains(&w) {
            let up = s.toggle(w);
            if w < s.f && w + s.m > s.f {
                assert_eq!(
                    up.g,
                    s.g + 1,
                    "removing atom generates gap for {i} and {w}"
                );
                assert!(up.apery_set.contains(&(w + s.m)));
                assert_eq!(up.apery_set[i], w + s.m);
                assert_eq!(up.r, s.r + 2, "r change w<f for i={i}");
                assert_eq!(up.sigma + 1, s.sigma);
            }
            if w == s.f + s.m {
                let rho = s.rho();
                assert_eq!(
                    up.g,
                    s.g + rho,
                    "removing atom generates rho gap for {i} and {w}"
                );
                assert!(up.apery_set.contains(&(w + rho * s.m)));
                assert_eq!(up.apery_set[i], w + rho * s.m);
                assert_eq!(up.r + rho * (s.m - 2), s.r, "r change w==f+m for i={i}");
                assert_eq!(
                    up.sigma,
                    rho * (s.m - 1) + s.sigma,
                    "sigma change w==f+m for i={i}"
                );
            }
        }
    }
}
\end{verbatim}
\section{Claude's review notes}
% inserted by claude — second pass, after fixes

\subsection*{What this writeup is about, in plain words}

You have a numerical semigroup $S$ with multiplicity $m$ and
Frobenius number $f$. Its Apéry set has one element $w_i$ per
residue class mod $m$, and the largest of these is
$w_\mu = f + m$ where $\mu = f \bmod m$. The writeup studies four
local \emph{moves} that turn $S$ into a nearby semigroup:

\begin{description}
    \item[$\mathcal{D}_{flip}$ (descent, flip):] Pick an Apéry
    element $w_i$ that lies in the open interval $(f, f+m)$ with
    $i \neq \mu$. Add $w_i - m$ to $S$. This fills in a reflected
    gap and lowers $w_i$ by $m$.
    \item[$\mathcal{D}_{total}$ (descent, total):] When the whole
    interval $V(S) = \{f-m+1,\dots,f-1\}$ already lies in $S$, add
    $f$ itself. This collapses the Apéry structure dramatically:
    the new Frobenius drops below $f-m$, many Apéry elements get
    re-indexed at once, and $\sigma$ decreases by $m-1$.
    \item[$\mathcal{A}_{flip}$ (ascent, flip):] The exact inverse
    of $\mathcal{D}_{flip}$: remove a generator $w_i \in V(S)$ with
    $i \neq \mu$.
    \item[$\mathcal{A}_{total}$ (ascent, total):] The harder one.
    Remove $w_\mu = f + m$ as a generator. Because $w_\mu$ is the
    only way to reach the next few rungs of its residue ladder, you
    actually lose $\rho$ elements in one move, where $\rho$ is the
    smallest $r_i$ over $i \neq \mu$.
\end{description}

The book-keeping (how $g$, $\sigma$, $r$ change in each case) is
mostly mechanical. What is interesting is whether you can
\emph{compose} these moves in a controlled way: take a descent
followed by an ascent and ask when you get back to the same
semigroup. That is the content of the ``Correspondence'' section.
The punchline is: a single descent--ascent pair is a round trip
only when the descent target has $\rho = 1$, i.e.\ the descent
lifts a semigroup with $w_\mu$ \emph{not} an atom to one with
$w_\mu$ a uniquely-decodable atom. For higher $\rho$, you can
iterate descents (provided $V(S) \subseteq S$ keeps holding), with
each step increasing $\rho$ by one and keeping $f + m$ an atom ---
until the chain hits a wall, namely a gap landing inside $V(S)$ or
$f$ dropping below $m$.

In short, this is a structural ``hill-climbing'' picture of the
descent/ascent dynamics on the Kunz/Apéry side of the Wilf
conjecture, with the four moves serving as elementary edges of an
operation graph on numerical semigroups.

\subsection*{Status of the issues from the first pass}

\begin{itemize}
    \item \textbf{Case~4 definition.} Fixed: now $\langle A(S) \setminus
    \{w_\mu\}\rangle$, with a paragraph clarifying why the
    naive set-minus is the wrong object.
    \item \textbf{$\rho > 0$ hypothesis in Case~4.} Fixed: stated
    inline as ``equivalently $\rho(S) > 0$'', and the equivalence
    is proved in the Correspondence section.
    \item \textbf{Index inconsistency $\mathcal{D}_{total}(i,\mu)$
    vs.\ $\mathcal{A}_{total}(i,m)$.} Fixed: the spurious second
    argument is gone everywhere; both families take only $\mu$.
    \item \textbf{Codomain of $D$.} Fixed: $D$ now lands in
    ``numerical semigroups with multiplicity $m$'' rather than
    $\mathbb{S}_\mu$. The old $\mathbb{S}_\mu$ has been renamed
    $\mathbb{S}_\mu^{\mathrm{old}}$ and serves only as the
    domain of $D$.
    \item \textbf{$r + \sigma = g$ justification.} Fixed: a
    one-sentence reason (no $\{x, f-x\} \subset S_+$ by closure,
    pair-counting on $[0,f]$) appears in the Setup section.
    \item \textbf{``Viewed from below'' equations.} Fixed: rewritten
    as $\rho$ descents undoing one ascent, with the matching
    arithmetic.
    \item \textbf{Minor wording / typos.} Fixed: ``necessesay /
    beeing / aways / The the / Semigroup'', stray standalone $S$,
    range of $L$ in RG, two Apéry definitions, ambiguous sum
    range, tautological ``minimal generator'' parenthetical, Wilf
    parenthetical.
\end{itemize}

\subsection*{Remaining issues (still open)}

\begin{enumerate}
    \item \textbf{No formal proofs.} The four change-formulas in
    Section~UpDown are stated and informally justified
    (``proved by tracking the Apéry set''). The codebase property
    test \verb|test_up_downs| checks them empirically over a wide
    fixture set, so they are very likely correct, but a reader of
    the writeup gets no algebraic derivation. Worth writing out at
    least one case (the $\sigma' = \sigma + \rho(m-1)$ case is the
    only one that surprises) as a small lemma.

    \item \textbf{$\rho(D(S)) = \rho(S) + 1$ asserted, not proved.}
    The Correspondence section still says ``the following
    properties are useful'' without supplying arguments. The
    \verb|s.r_i(i)| identity for Kunz coefficients is the natural
    proof tool, but the writeup doesn't reference it.

    \item \textbf{$V(D(S)) \subseteq D(S)$ termination condition.}
    The iterated descent chain is described informally as
    ``stops once there is a gap in $V$ or $f < m$''. There is
    presumably a clean characterisation of which semigroups admit
    a chain of length $\ge k$, but the writeup doesn't attempt one.
    Could be stated either combinatorially (in terms of the Apéry
    set) or via the Kunz matrix.

    \item \textbf{Symmetry / direction conventions.} ``Descent''
    \emph{adds} elements to $S$; ``ascent'' \emph{removes} them.
    This convention (descend = grow) is consistent with the
    codebase but is the reverse of what some readers might expect
    from the words. A one-line note up front (``descent shrinks
    the gap set; ascent enlarges it'') would help.

    \item \textbf{Wilf-aside is a teaser, not a result.} The line
    ``we really should work out the change in $e$'' is a TODO; no
    statement about how $e$ moves under any of the four operations
    is given. If the writeup is meant to be self-contained even as
    an informal note, this is the most natural next section.

    \item \textbf{Code listing is not commented on.} The verbatim
    Rust test in Section~Code is given without text explaining
    which assertion corresponds to which case in
    Section~UpDown. Adding case labels in comments
    (\verb|// Case 1: D_flip|, etc.) would help the reader who
    wants to cross-check the math against the code.

    \item \textbf{Where $\mathcal{D}_{flip}$ / $\mathcal{A}_{flip}$
    live in the descent graph.} Section~Correspondence treats only
    the \emph{total} pair. Whether the flip pair admits any
    analogous round-trip discussion is not addressed. The
    inversion remark ``$\mathcal{A}_{flip}$ inverts
    $\mathcal{D}_{flip}$'' is a single sentence; the consequence
    (that the flip pair is always a round-trip on the appropriate
    domain) deserves a paragraph.

    \item \textbf{Notation drift between $r_i$ and $c(i,j)$.}
    Internally to the codebase, $r_i = c(i, \mu - i)$ where
    $c(\cdot,\cdot)$ is the Kunz coefficient. The writeup never
    introduces $c$, even though the Kunz interpretation makes some
    arguments (especially $\rho(D(S)) = \rho(S) + 1$) crisper.
\end{enumerate}

\subsection*{Notation $\leftrightarrow$ codebase}

\begin{center}
	\begin{tabular}{l|l}
		LaTeX & Code \\\hline
		$m,\,f,\,\mu$           & \verb|s.m|, \verb|s.f|, \verb|s.mu| \\
		$g,\,\sigma,\,r$        & \verb|s.g|, \verb|s.sigma|, \verb|s.r| \\
		$r_i,\,\rho$            & \verb|s.r_i(i)|, \verb|s.rho()| \\
		$w_i,\,\operatorname{Ap}(S)$ & \verb|s.apery_set[i]|, \verb|s.apery_set| \\
		$\operatorname{RG}(S)$  & \verb|s.blob()| \\
		$V(S) \subseteq S$      & \verb|s.v_in_s()| \\
		atom                    & element of \verb|s.gen_set| \\
		$S^\downarrow$ (descent) & \verb|s.toggle(w - s.m)| \\
		$S^\uparrow$ (ascent)    & \verb|s.toggle(w)| \\
		$t$                     & \verb|s.t| (Rust) / \verb|s.type_t| (JS)
	\end{tabular}
\end{center}


\end{document}
